---
title: "Simulating Data from a Linear Model"
subtitle: "Creating a synthetic continuous-outcome dataset with a known Gaussian generating process"
---

This chapter creates a synthetic dataset in which the outcome is continuous and generated by a linear model with additive Normal noise. The design is motivated by blood-pressure treatment studies such as SPRINT, where a continuous clinical measurement is observed after treatment and analyzed in relation to baseline risk factors and treatment assignment @sprint2015. The synthetic version here is not a replication of SPRINT. Instead, it creates a simplified follow-up systolic blood pressure outcome so that the full data-generating process is known in advance. That makes it useful for checking whether an ordinary least squares regression can recover the true coefficients.

The practical reason to simulate linear-model data is straightforward. Continuous outcomes appear across health economics and decision sciences: systolic blood pressure, quality-of-life scores, biomarker levels, total costs after transformation, recovery time, and resource use summaries. If the conditional mean is approximately linear and the residual variation is additive, the linear model is often the first analytic tool to test. Simulation makes that structure explicit.

## What variables will be created

The synthetic sample will represent patients followed for 6-month systolic blood pressure after treatment intensification. `baseline_sbp` will measure baseline systolic blood pressure in mmHg. `age` will represent age in years. `chronic` will count the number of chronic conditions. `intensive_treatment` will indicate whether the patient was assigned to a more intensive management strategy. `adherence` will indicate whether the patient remained adherent to the assigned regimen. The outcome `followup_sbp` will measure systolic blood pressure after 6 months.

These variables give the linear model something realistic to work with: a continuous baseline predictor, a treatment indicator, a comorbidity burden, and a behavioral factor that shifts the follow-up mean.

## The data-generating process

The core assumption is that the continuous outcome follows

$$
Y_i = \beta_0 +
\beta_1 \text{baseline\_sbp}_i +
\beta_2 \text{intensive\_treatment}_i +
\beta_3 \text{age}_i +
\beta_4 \text{chronic}_i +
\beta_5 \text{adherence}_i +
\varepsilon_i,
$$

where the residual term is Normally distributed:

$$
\varepsilon_i \sim \mathcal{N}(0, \sigma^2).
$$

For this simulation, the true parameters are set to

$$
\beta_0 = 35,\;
\beta_1 = 0.62,\;
\beta_2 = -7.5,\;
\beta_3 = 0.18,\;
\beta_4 = 1.6,\;
\beta_5 = -4.5,
$$

with

$$
\sigma = 8.
$$

These values imply a clinically sensible pattern. Higher baseline blood pressure predicts higher follow-up blood pressure. Intensive treatment lowers follow-up systolic pressure. Older age and more chronic conditions raise it slightly. Better adherence lowers it further.

## Step 1: Generate the synthetic sample

```{r}
set.seed(2026)

n <- 6000

age <- pmax(round(rnorm(n, mean = 64, sd = 9)), 40)
chronic <- pmin(rpois(n, lambda = 2.1), 7)
intensive_treatment <- rbinom(n, size = 1, prob = 0.5)

baseline_sbp <- round(
  rnorm(
    n,
    mean = 142 + 0.45 * chronic + 0.18 * (age - 64),
    sd = 11
  ),
  1
)

adherence <- rbinom(
  n,
  size = 1,
  prob = plogis(0.9 + 0.35 * intensive_treatment - 0.22 * chronic - 0.02 * (age - 64))
)

true_mean <- 35 +
  0.62 * baseline_sbp -
  7.5 * intensive_treatment +
  0.18 * age +
  1.6 * chronic -
  4.5 * adherence

followup_sbp <- true_mean + rnorm(n, mean = 0, sd = 8)

synthetic_bp <- data.frame(
  followup_sbp,
  baseline_sbp,
  age,
  chronic,
  intensive_treatment,
  adherence,
  true_mean
)

simulation_summary <- data.frame(
  sample_size = nrow(synthetic_bp),
  mean_followup_sbp = mean(synthetic_bp$followup_sbp),
  mean_baseline_sbp = mean(synthetic_bp$baseline_sbp),
  mean_age = mean(synthetic_bp$age),
  mean_chronic = mean(synthetic_bp$chronic),
  adherence_rate = mean(synthetic_bp$adherence)
)

simulation_summary[, -1] <- round(simulation_summary[, -1], 3)

knitr::kable(
  simulation_summary,
  caption = "Summary of the synthetic linear-model dataset"
)
```

That code first builds the predictors, then uses the linear predictor to generate the expected follow-up blood pressure, and finally adds Gaussian noise around that expectation. The `true_mean` column is especially useful because it stores the deterministic part of the generating process before residual error is added.

## Step 2: Fit the model that matches the truth

Now fit the same linear model used to generate the data:

```{r}
linear_fit <- lm(
  followup_sbp ~ baseline_sbp + intensive_treatment + age + chronic + adherence,
  data = synthetic_bp
)

truth <- c(
  "(Intercept)" = 35,
  baseline_sbp = 0.62,
  intensive_treatment = -7.5,
  age = 0.18,
  chronic = 1.6,
  adherence = -4.5
)

comparison_table <- data.frame(
  term = names(coef(linear_fit)),
  true_value = truth[names(coef(linear_fit))],
  estimated_value = coef(linear_fit)
)

comparison_table$bias <- comparison_table$estimated_value - comparison_table$true_value
comparison_table[, c("true_value", "estimated_value", "bias")] <-
  round(comparison_table[, c("true_value", "estimated_value", "bias")], 3)

knitr::kable(
  comparison_table,
  caption = "True and estimated coefficients under the correctly specified linear model"
)
```

This is the main value of simulation. Because the truth is known, coefficient recovery can be checked directly. Under correct specification and a reasonably large sample, the fitted estimates should be close to the generating values.

## Step 3: Compare the residual standard deviation with the truth

The error variance is part of the generating process too, so it is worth checking whether the fitted model recovers it.

```{r}
sigma_table <- data.frame(
  quantity = c(
    "True residual standard deviation",
    "Estimated residual standard deviation",
    "True residual variance",
    "Estimated residual variance",
    "R-squared"
  ),
  value = c(
    8,
    summary(linear_fit)$sigma,
    8^2,
    summary(linear_fit)$sigma^2,
    summary(linear_fit)$r.squared
  )
)

sigma_table$value <- round(sigma_table$value, 3)

knitr::kable(
  sigma_table,
  caption = "Recovery of the residual variation in the synthetic linear model"
)
```

The residual standard deviation should not match the true value exactly in every sample, but it should be close when the sample is large and the model is correctly specified.

## Step 4: Compare true and fitted conditional means across treatment profiles

Coefficients are useful, but fitted means are easier to interpret. The next block compares the true and fitted follow-up systolic pressure across a range of baseline blood pressure values for the two treatment groups.

```{r}
profiles <- expand.grid(
  baseline_sbp = seq(120, 180, by = 2),
  intensive_treatment = c(0, 1)
)

profiles$age <- 64
profiles$chronic <- 2
profiles$adherence <- 1

profiles$true_mean <- 35 +
  0.62 * profiles$baseline_sbp -
  7.5 * profiles$intensive_treatment +
  0.18 * profiles$age +
  1.6 * profiles$chronic -
  4.5 * profiles$adherence

profiles$fitted_mean <- predict(linear_fit, newdata = profiles)
profiles$treatment_group <- ifelse(
  profiles$intensive_treatment == 1,
  "Intensive treatment",
  "Standard treatment"
)

ggplot2::ggplot(
  profiles,
  ggplot2::aes(x = baseline_sbp, color = treatment_group)
) +
  ggplot2::geom_line(
    ggplot2::aes(y = true_mean),
    linewidth = 1.1
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = fitted_mean),
    linetype = 2,
    linewidth = 0.9
  ) +
  ggplot2::labs(
    title = "True and fitted mean follow-up blood pressure in the synthetic linear dataset",
    subtitle = "Solid lines show the generating means; dashed lines show the fitted OLS model",
    x = "Baseline systolic blood pressure",
    y = "Expected follow-up systolic blood pressure",
    color = "Profile"
  ) +
  ggplot2::scale_color_manual(values = c("#0b5d4b", "#b54708")) +
  ggplot2::theme_minimal(base_size = 12)
```

If the dashed lines stay close to the solid lines, the fitted linear model is reproducing the conditional mean function well.

## Step 5: Compare observed and true means by risk subgroup

One more useful check is to compare average observed outcomes with the true generating means across simple subgroups.

```{r}
synthetic_bp$risk_group <- cut(
  synthetic_bp$baseline_sbp,
  breaks = c(-Inf, 130, 145, 160, Inf),
  labels = c("<=130", "131-145", "146-160", ">160")
)

subgroup_table <- aggregate(
  cbind(followup_sbp, true_mean) ~ risk_group + intensive_treatment,
  data = synthetic_bp,
  mean
)

subgroup_table$intensive_treatment <- ifelse(
  subgroup_table$intensive_treatment == 1,
  "Intensive",
  "Standard"
)

subgroup_table[, c("followup_sbp", "true_mean")] <-
  round(subgroup_table[, c("followup_sbp", "true_mean")], 2)

knitr::kable(
  subgroup_table,
  caption = "Observed and generating subgroup means in the synthetic dataset"
)
```

This table is another reminder that simulation creates two different quantities: the true conditional mean and the realized sample outcome. The observed mean is noisy. The true mean is not.

## Main assumptions behind this simulation

The most important assumption is linearity in the conditional mean:

$$
\mathbb{E}(Y_i \mid X_i) = X_i \beta.
$$

The second assumption is homoskedastic Gaussian noise:

$$
\varepsilon_i \sim \mathcal{N}(0, \sigma^2).
$$

The third is conditional independence across observations. Those assumptions are useful for learning because they create a clean benchmark in which ordinary least squares is correctly specified.

Real data are often more complicated. Effects may be nonlinear, residual variance may increase with baseline risk, treatment assignment may be confounded, and outcomes may be censored or bounded. That is exactly why simulation is valuable. It gives a controlled environment before those complications are added.

## How to adapt this template

Once the basic structure is clear, the same template can be modified easily. You can add interactions, create nonlinear baseline effects, introduce heteroskedasticity, correlate treatment assignment with baseline risk, or add clustered errors. Each change creates a new synthetic laboratory for studying how a fitted linear model behaves under a different truth.

In practice, this is often the most useful way to learn what ordinary least squares is actually doing. How biased does the treatment coefficient become when an important covariate is omitted? How much sample size is needed before a small treatment effect is recovered reliably? What happens when the true relationship is quadratic but the fitted model is linear? Synthetic data can answer those questions directly.

## Further reading

The linear-model framework remains a central entry point for applied regression analysis, and blood-pressure trials such as SPRINT provide a useful substantive anchor for continuous clinical outcomes @sprint2015. A natural next step after this chapter is to extend the simulation by adding heteroskedasticity, treatment-effect heterogeneity, or clustering and then studying how coefficient recovery changes under misspecification.
